Circularity in Population-Based FA Analysis

### 2011-06-08 ANTsTBSS/paper.tex
First draft submission.

%\documentclass[preprint,authoryear,review,12pt]{elsarticle}
\documentclass[final,5p,times,twocolumn]{elsarticle}

@@

\title{On Population-Based FA Analysis with Non-independent Normalization and Statistics}

@@

\begin{abstract}
Template-based normalization plays a critical role in many population
studies based on diffusion tensor imaging (DTI).  Several current
approaches compute the fractional anisotropy (FA) from the DTI and then
directly register the FA image to a FA template before performing
statistical comparison between different groups within a population.  In
this paper, we demonstrate that this analysis strategy induces selection
bias that varies in strength with the similarity metric used in
registration.  Specifically, we show that the popular sum of squared
difference intensity metric maximizes effect size, rather than
anatomical alignment, and consistently overestimates statistical
significance.  Consequently, we advocate a more traditional approach
that normalizes a population of FA images through the associated 
anatomical images.  We show that the Symmetric Group Normalization
strategy for unbiased template construction from anatomical (e.g.
T1-weighted) images provides a valid and unbiased coordinate system for
statistical inference of FA differences and yet retains detection power
in comparison to FA-based normalization methods.  We establish this
outcome by using open data comprised of normal controls from the International
Neuroimaging Data-sharing Initiative as well as a cohort of 16 traumatic
brain injury survivors compared with 17 control subjects.  Alignment
comparisons with the normalization component of the popular tract-based
spatial statistics (TBSS) show that our anatomically-based template
approach produces good alignments despite the absence of any direct
FA-to-FA registration.  Of practical consideration we note that the
software tools discussed in this work are provided to the public as part
of the open source Advanced Normalization Tools (ANTs) repository.  In conclusion, we recommend against circular analysis strategies that select image similarity metrics that are tied directly to the ultimate statistical functions used to assess population effects.

@@

\begin{keyword}
Advanced Normalization Tools (ANTs) \sep diffusion tensor imaging \sep fractional anisotropy \sep template construction \sep traumatic brain injury
%% keywords here, in the form: keyword \sep keyword
\end{keyword}

@@

The seminal work of Basser et al. \cite{Basser1994a,Basser1994} established diffusion tensor imaging (DTI) as a viable investigatory MRI technique.  DTI's sensitivity to brain architecture \cite{Basser1996,Assaf2008} enables multiple uses in assessing neuro-structural differences in cross-population studies \cite{Kubicki2005,Arnone2008,Kantarci2010,Rametti2010} including subjects with traumatic brain injury (TBI) \cite{Caeyenberghs2010,Jiang2010,Warner2010}.  Two popular comprehensive software packages for assessing population FA differences include the SPM Matlab-based toolkit%
\footnote{
http://www.fil.ion.ucl.ac.uk/spm/
}
for voxel-based morphometry (VBM)
and the tract-based spatial statistics (TBSS) framework \cite{Smith2006}.%
\footnote{
http://www.fmrib.ox.ac.uk/fsl/tbss/index.html
} 
VBM \cite{Ashburner2001} continues to find application to FA studies \cite{Kakeda2010,Takao2010} despite concerns about its general validity \cite{Bookstein2001,Davatzikos2004} and with respect to DTI \cite{Jones2005,Chung2008}.  The TBSS framework was partially developed in response to these concerns in which voxel values are projected onto a template white matter skeleton for increased statistical power.  
%(e.g. \cite{Arnone2008,Bodini2009}).

Although there are substantive differences in normalization
between these and other analysis approaches, a core
commonality includes direct FA-to-FA registration.  The sum of squared differences (SSD) image similarity metric is perhaps the easiest to interpret (it drives the image difference to zero), is computationally efficient (just compute the image difference and gradient) and is therefore widely used.  Some of the most popular registration methods (SPM, Demons, FSL's FNIRT) rely on this metric and yield reasonable performance levels \cite{Klein2009}.%
\footnote{
In addition to TBSS and SPM, other popular packages in the neuroimaging community 
which use the SSD metric include AIR \cite{Woods1998} and LDDMM \cite{Beg2005}.
}

@@

Thus, instead of normalization based on anatomical alignment followed by
statistical testing, direct FA normalization explicitly conflates
anatomical alignment with optimizing the very statistical testing
results that one will ultimately use to assess hypotheses.  This immediately evokes a sense of circularity in the analysis \cite{Kriegeskorte2010}.
According to Kriegeskrote et al., ``An analysis is circular (or nonindependent) if it is based on data that were selected for showing the effect of interest or a related effect.''  As illustrated with
Eqn. \ref{eq:variance}, direct FA-to-FA registration will produce transformations that align (or select) voxels such that the statistical testing result (or the effect of interest) is increased (assuming one is using Student's 
$t$-test).  While the SSD metric is able to provide visually attractive alignments across subjects, the voxel-level details (that is, which voxels are matched where) are, as we show in the following simulation study, strikingly important in terms of the effect on detection power.

@@

\begin{figure}
\begin{tabular}{c}
  \includegraphics[width=85mm]{simulatedData.jpg}
\end{tabular}
\caption{(a) Axial slice from the discrete BrainWeb data set.  From (a) 
we create the (b) simulated template and sample (c) control and (d) patient images with both 
additive noise and subsequent Gaussian smoothing. }
\label{fig:simulated_data}
\end{figure}

\begin{figure}
\begin{tabular}{c}
  \includegraphics[width=85mm]{simulatedDataWarped.jpg}
\end{tabular}
\caption{Warping the image in Fig. \ref{fig:simulated_data}(c) to the template in 
  Fig. \ref{fig:simulated_data}(b) using the SSD similarity metric produces the image shown in
  (a).  The corresponding deformation is given in (b).  Although the images were anatomically
  aligned prior to invoking ANTs registration, deformation is driven by the similarity metric
  causing the accumulation of Type 1 errors.}
\label{fig:simulated_data_warped}
\end{figure}

Axial slice number 159 was extracted from \verb#subject_04# of the discrete BrainWeb database
\cite{Aubert-Broche2006} which is shown in Fig. \ref{fig:simulated_data}.  After thresholding out 
the white matter and assigning all white matter voxels a value of 0.7, 10\% of these white 
matter voxels were randomly selected to be assigned a value of 0.6 in the simulated control 
template and a value of 0.3 in the simulated patient template.  
Note that these values were arbitrarily selected to be more or less in the range of typical
FA values.  Gaussian noise (mean = 0.0, standard deviation = 0.1) was added to the white
matter template followed by a Gaussian smoothing (standard deviation = 1.5 mm) shown in Fig. \ref{fig:simulated_data}(b).  A simulated control cohort was created consisting of 20 subjects.  
Starting with the simulated control template, each of the 20 subjects is created by adding
Gaussian noise (mean = 0.0, standard deviation = 0.1) followed by 
Gaussian smoothing (standard deviation = 1.5 mm).  Thus, each subject is differentiated only by a unique random draw from the noise distribution and there is no real shape difference between subjects.  A simulated patient cohort of 20 subjects
was created similarly using the simulated patient template.  A sample control subject and sample
patient subject are given, respectively, in Fig. \ref{fig:simulated_data}(c) and (d).

\begin{figure*}
\begin{center}
\begin{tabular}{c}
  \includegraphics[width=180mm]{simulatedTesting.jpg}
\end{tabular}
\caption{Student's $t$-test and the nonparametric Wilcoxon rank-sum test results (one-tailed, $\alpha = 0.95$) for the simulated experiment adjusted for multiple comparisons (FDR).
The true significance curves are drawn in black.  Elevated statistical power is seen with all 4 
         metrics for both tests although the effect is less in the more robust Wilcoxon results. Not shown are the significance curves for the Brunner-Munzel testing \cite{Rorden2007} which showed similar effects to the results on the left.
        }
\label{fig:simulated_testing}
\end{center}        
\end{figure*}

@@

To illustrate the effect of similarity metric on statistical significance, we performed ANTs registration \cite{Avants2011} of each subject to the template
(Fig. \ref{fig:simulated_data}(b)) using cross correlation (CC), mutual information (MI), 
Demons \cite{Thirion1998}, and sum of squared differences (SSD). A representative deformed image
using the SSD metric and its corresponding deformed grid is given in 
Fig. \ref{fig:simulated_data_warped}. 
The resulting statistical significance curves are plotted with the true significance curves in Fig. 
\ref{fig:simulated_testing}.

@@

Common similarity metric functions can therefore induce false positives by minimizing local intensity 
variance between the subjects and template which explicitly maximizes statistical
testing results in contrast to maximizing anatomical alignment.  This effect is strongest in 
the SSD and Demons metrics and weakest in the MI and CC metrics likely due
to the degree of correspondence to Eqn. (\ref{eq:variance}).   This is also supported by contrasting
the $t$-test results given in Fig. \ref{fig:simulated_testing}(a) with the non-parametric Wilcoxon
rank-sum testing (Fig. \ref{fig:simulated_testing}(b)).  The rank test does not explicitly use variance in its assessment and thus exhibits far less statistical exaggeration even though it offers roughly 95\%
the power of the $t$-test assuming certain assumptions are met \cite{Siegel1988,Rorden2007}.

Although various processing choices mitigate these effects, such as Gaussian smoothing, nonparametric testing, and the use of more robust similarity metrics, such choices are often made on an ad hoc basis.  
In contrast, in this work we strongly advocate the general application
%A more principled approach is the use of an anatomical, i.e. T1-weighted, unbiased shape/intensity 
%template constructed from the population images.  DTI-derived scalar images can then be mapped 
%to their corresponding T1-weighted image in a highly constrained manner and subsequently 
%mapped to the template via nonrigid registration of the T1-weighted image.  This minimizes 
%possible complications associated with direct FA-to-FA image registration.
%We recognize that using the anatomical images in the alignment process has been previously proposed (in fact, such a suggestion occurs in the {\it Future directions} section of the original TBSS paper).  
of a high performance \cite{Klein2009}, nonrigid registration algorithm \cite{Avants2011}, which forms the underlying machinery for an optimal template construction algorithm \cite{Avants2010}, in facilitating alignment between DTI-derived scalar images for cross-population studies.  The key point, here, is that the FA images are mapped to the common reference space via intra-subject DT-to-T1 mapping followed by T1-to-T1 alignment.  The T1-to-T1 step is (statistically) independent of the FA values which are the ultimate statistical target of hypothesis testing in FA population studies.  Below, we show that this same pattern of effects is reproduced in real data and detail our proposed solution.

\begin{figure*}
\begin{center}
\begin{tabular}{c}
  \includegraphics[width=170mm]{workflow.jpg}
\end{tabular}
\caption{Possible ANTs-based TBSS workflow using SyGN.  In contrast to standard TBSS in which an individual subject's FA is registered to the FMRIB58 template
(transform represented with the blue arrow), our proposed modification first constructs an ANTs optimal anatomical template as illustrated in the center of the figure.  Each FA image is then rigidly mapped, with distortion correction, to the individual T1 using the white matter mask as shown on the left.  The individual T1-weighted image is then nonrigidly registered to the anatomical template.  Optionally, one can register the T1 template to the MNI152 template which resides in the same space as the FMRIB58 template.  Thus, the composite transform, carefully constructed without any FA-to-FA image registration, can take an individual subject's FA map (or other DTI-derived measure) to the desired standardized coordinate system.    }
\label{fig:workflow}
\end{center}
\end{figure*}

@@

%
%
%
%It was precisely due to these concerns that the TBSS framework was developed constituting an entire processing pipeline spanning from the reconstruction of the DTI tensorial data
%%to the calculation of the corresponding DTI-derived scalar measures (typically fractional anisotropy (FA)), to the nonrigid alignment of the population to a standard coordinate system and skeletonization, and, finally, 
%through to the statistical analysis for discerning differences across populations.  Major intermediate steps include the calculation of the corresponding DTI-derived scalar measures (typically fractional anisotropy (FA) \cite{Basser1996}), nonrigid alignment of the population to a standard coordinate system, and subsequent ``skeletonization'' of the mean of the aligned FA maps.  The alignment to a standard space is performed in such a way to remove the smoothing step associated with SPM.  Skeletonization and its projection across all registered DTI-derived scalar images is meant to accommodate the possibility of imperfectly aligned data.  However, as shown in a recent study \cite{Zalesky2011}, although the alignment error is reduced with the skeleton projection step, misalignments still remain post-projection.  Added to this is the potential confounding effect of minimizing data differences via nonrigid registration which are potentially the precise differences to be statistically quantified.

@@

%A crucial component of TBSS is the alignment of the DTI-derived scalar images \cite{Zalesky2011}. 
%
%Voxel-based morphometry vs. tractography vs. tract-specific analysis
%
%Talk about {\em skeleton projection} and how our work is different from \cite{Zalesky2011}.
%Also talk about the advantages over building a dti-based template as discussed in 
%\cite{Hecke2010}.

@@

\subsection{Anatomical Image Template Construction}
In general, we normalize images to to a standard coordinate system to
reduce intersubject variability in population studies.  Various
approaches exist for determining the normalized space such as selection
of a pre-existing template based on a single subject, e.g. the Talairach
atlas \cite{Talairach1988}), or a publicly available averaged group of
subjects, e.g. the MNI \cite{Collins1994} or ICBM \cite{Mazziotta1995}
templates.  One challenge with standard templates is that they may inadvertantly bias one's results by enabling better normalization of subjects to which the template is more similar.  This issue is exacerbated when dealing with populations that have high variance (e.g. due to disease) and/or when one's normalization method is low-dimensional (not flexible enough to capture large shape differences).

Population-specific templates (or optimal templates) alleviate some of
these issues by deriving a most representative image from the population
\cite{Good2001}.  Large deformation registration algorithms also reduce
this confound by being less sensitive to the deformation distance
between subject and target.  Some approaches combine both advantages,
for instance, the diffeomorphic approach of Joshi et al. employs the
SSD metric and a shape distance to bring the subject group of images
into alignment \cite{Joshi2004}.  Variants
include extension to multiple modalities \cite{Lorenzen2006} and small deformations
\cite{Geng2009}.  These approaches iteratively minimize group difference in ``congealing''
towards a representative image template \cite{Learned-Miller2006}.

In contrast, Avants et al. explicitly model the geometric component of the 
normalized space during optimization.  Coupling the intrinsic symmetry of 
SyN pairwise registration \cite{Avants2011} and an
optimized sharpening/averaging of the template appearance, Symmetric Group
Normalization (SyGN) is a powerful framework for producing optimal population-specific
templates \cite{Avants2010} with arbitrary similarity metric choice. Given a set of representative images, 
$\{I_1, I_2, \ldots, I_M\}$, optimization involves finding the set of paired
diffeomorphic transformations, $\left\{\left(\phi^1_1, \phi^1_2\right), 
\left(\phi^2_1, \phi^2_2\right), \ldots, \left(\phi^M_1, \phi^M_2\right) \right\}$,
the optimal template appearance, $J^*$, and corresponding coordinate system, $\psi(\mathbf{x})$,
which minimize the following cost function:
\begin{align}
  \sum_{m=1}^M \left[
    D\left( \psi(\mathbf{x}), \phi^m_1(\mathbf{x}, 1) \right) + 
    \Pi \left( I_m\left(\phi^m_2(\mathbf{x}, 0.5) \right), J^*\left(\phi^m_1(\mathbf{x}, 0.5) \right) \right)
    \right]
\end{align}
where $D$ is the diffeomorphic shape distance, 
\begin{align}
  D\left(\phi(\mathbf{x}, 0), \phi(\mathbf{x}, 1)\right) = \int_{0}^1 \| v(\mathbf{x}, t) \|_L dt, 
\end{align}
dependent upon the choice of the linear operator, $L$, and
$v$ is the diffeomorphism-generating velocity field, 
\begin{align}
  v\left(\phi(\mathbf{x}, t), t \right) = \frac{d\phi(\mathbf{x}, t)}{dt},\,\,\, \phi(\mathbf{x}, 0) = \mathbf{x}.
\end{align}
$\Pi$ is the choice of similarity metric, often cross-correlation \cite{Avants2008a}, calculated in the 
virtual domain midway between each individual image and the current estimate of the template.

@@

\begin{enumerate}
  \item Build the anatomical template from  the population sample or sample subset using SyGN as described in
  the previous section.
  \item If only using a sample subset for template creation, register the remaining
        anatomical data to the template and save the resulting transformation for each         
        subject indexed by $i$ (i.e. $\mathcal{T}^i_{template}$).
  \item Find the optimal transformation from each
        subject's FA to the anatomical template:
  \begin{enumerate}
  \item For each anatomical image, segment the white matter (in ANTs, one can use N4 bias correction \cite{Tustison2010} and Atropos $n$-tissue segmentation \cite{Avants2011a}).
  \item Rigidly transform each subject's average DWI to the bias-corrected anatomical image ($\mathcal{R}^i_{dwi}$).
  \item Perform distortion correction on the FA to anatomical mapping 
        by localized nonrigid registration
        of rigidly transformed FA map ($=\mathcal{R}^i_{dwi} \mathrm{FA}$) using
        the white matter mask obtained in step 3(a) ($\mathcal{T}^i_{FA}$).
  \item For each subject, concatenate the transforms which maps each
        subject's FA to the template ($\mathcal{T}^i_{total} = \mathcal{T}^i_{template} \circ 
        \mathcal{T}^i_{FA} \circ \mathcal{R}^i_{dwi} \mathrm{FA}$).
  \item Warp each subject's FA map to the template using $\mathcal{T}^i_{total}$.  
  Equivalently, one could warp the mean diffusion (MD), axial diffusion (AD) or any 
  other scalar derivable from DTI to the template.  
%  One other detail is that we prefer to warp the subject's DTI to 
%  the template and then, in template space, derive the scalar images of interest.  
  \end{enumerate}      
%  \item Perform the remaining portions of the standard TBSS pipeline:
%    \begin{enumerate}
%      \item Create the mean FA image from all template-aligned FA images.
%      \item Produce the white matter skeleton from the mean FA image.
%      \item Create skeleton distance map and project all FA data onto the skeleton.
%      \item Run the statistical analysis.
%    \end{enumerate}
\end{enumerate}
The next section compares this framework to FA-based normalization strategies with respect to both a patient-control study and a study of gender and age effects in a control population.

>>>>

\documentclass[preprint,authoryear,review,12pt]{elsarticle}
%\documentclass[final,5p,times,twocolumn]{elsarticle}

@@

\title{Circularity in Population-Based FA Analysis}

@@

\begin{abstract}
Template-based normalization plays a critical role in many population
studies based on diffusion tensor imaging (DTI).  Several current
approaches compute the fractional anisotropy (FA) from the DTI and then
directly register the FA image to a FA template before performing
statistical comparison between different groups within a population.  In
this paper, we demonstrate that this analysis strategy induces selection
bias that varies in strength with the similarity metric used in
registration.  Specifically, we show that the popular sum of squared
difference intensity metric maximizes effect size, rather than
anatomical alignment, and consistently overestimates statistical
significance.  Consequently, we advocate a more cautious approach
that normalizes a population of FA images through the associated 
anatomical images.  We show that the Symmetric Group Normalization
strategy for unbiased template construction from anatomical (e.g.
T1-weighted) images provides a valid and unbiased coordinate system for
statistical inference of FA differences and yet retains detection power
in comparison to FA-based normalization methods.  We establish this
outcome by using open data comprised of normal controls from the International
Neuroimaging Data-sharing Initiative as well as a cohort of 16 traumatic
brain injury survivors compared with 17 control subjects.  Alignment
comparisons with the normalization component of the popular tract-based
spatial statistics (TBSS) show that our anatomically-based template
approach produces good alignments despite the absence of any direct
FA-to-FA registration.  Of practical consideration we note that the
software tools discussed in this work are provided to the public as part
of the open source Advanced Normalization Tools (ANTs) repository.  In conclusion, we recommend against circular analysis strategies that select image similarity metrics that are tied directly to the statistical functions used to assess population effects.

@@

\begin{keyword}
Advanced Normalization Tools (ANTs) \sep diffusion tensor imaging \sep fractional anisotropy \sep template construction
%% keywords here, in the form: keyword \sep keyword
\end{keyword}

@@

The seminal work of Basser et al. \citep{Basser1994a,Basser1994} established diffusion tensor imaging (DTI) as a viable investigatory MRI technique.  DTI's sensitivity to brain architecture \citep{Basser1996,Assaf2008} enables promising analysis possibilities assessing neuro-structural differences in cross-population studies \citep{Kubicki2005,Arnone2008,Kantarci2010,Rametti2010} including subjects with traumatic brain injury (TBI) \citep{Caeyenberghs2010,Jiang2010,Warner2010}.  Two popular comprehensive software packages for assessing population FA differences include the SPM Matlab-based toolkit%
\footnote{
http://www.fil.ion.ucl.ac.uk/spm/
}
for voxel-based morphometry (VBM)
and the tract-based spatial statistics (TBSS) framework \citep{Smith2006}.%
\footnote{
http://www.fmrib.ox.ac.uk/fsl/tbss/index.html
} 
VBM \citep{Ashburner2001} continues to find application to FA studies \citep{Kakeda2010,Takao2010} despite concerns about its general validity \citep{Bookstein2001,Davatzikos2004} and with respect to DTI \citep{Jones2005,Chung2008}.  The TBSS framework was partially developed in response to these concerns in which voxel values are projected onto a template white matter skeleton for increased statistical power.  
%(e.g. \citep{Arnone2008,Bodini2009}).

Although there are substantive differences in normalization
between these and other analysis approaches, a core
commonality includes direct FA-to-FA registration.  The sum of squared differences (SSD) image similarity metric is perhaps the easiest to interpret (it drives the image difference to zero), is computationally efficient (just compute the image difference and gradient) and is therefore widely used.  Some of the most popular registration methods (SPM, Demons, FSL's FNIRT) rely on this metric and yield reasonable performance levels \citep{Klein2009}.%
\footnote{
In addition to TBSS and SPM, other popular packages in the neuroimaging community 
which use the SSD metric include AIR \citep{Woods1998} and LDDMM \citep{Beg2005}.
}

@@

Thus, instead of normalization based on anatomical alignment followed by
statistical testing, direct FA normalization explicitly conflates
anatomical alignment with optimizing the very statistical testing
results that one will ultimately use to assess hypotheses.  This immediately evokes a sense of circularity in the analysis \citep{Kriegeskorte2010}.
According to Kriegeskrote et al., ``An analysis is circular (or nonindependent) if it is based on data that were selected for showing the effect of interest or a related effect.''  As illustrated with
Eqn. \ref{eq:variance}, direct FA-to-FA registration will produce transformations that align (or select) voxels such that the statistical testing result (or the effect of interest) is increased (assuming one is using Student's 
$t$-test).  While the SSD metric is able to provide visually attractive alignments across subjects, the voxel-level details (that is, which voxels are matched where) are, as we show in the following simulation study, strikingly important in terms of the effect on detection power.

@@

\begin{figure}
\begin{center}
\begin{tabular}{c}
  \includegraphics[width=85mm]{simulatedData.pdf}
\end{tabular}
\caption{(a) Axial slice from the discrete BrainWeb data set.  From (a) 
we create the (b) simulated template and sample (c) control and (d) patient images with both 
additive noise and subsequent Gaussian smoothing. }
\label{fig:simulated_data}
\end{center}
\end{figure}

\begin{figure}
\begin{center}
\begin{tabular}{c}
  \includegraphics[width=85mm]{simulatedDataWarped.pdf}
\end{tabular}
\caption{Warping the image in Fig. \ref{fig:simulated_data}(c) to the template in 
  Fig. \ref{fig:simulated_data}(b) using the SSD similarity metric produces the image shown in
  (a).  The corresponding deformation is given in (b).  Although the images were anatomically
  aligned prior to invoking ANTs registration, deformation is driven by the similarity metric
  causing the accumulation of Type 1 errors.}
\label{fig:simulated_data_warped}
\end{center}
\end{figure}

Axial slice number 159 was extracted from \verb#subject_04# of the discrete BrainWeb database
\citep{Aubert-Broche2006} which is shown in Fig. \ref{fig:simulated_data}.  After thresholding out 
the white matter and assigning all white matter voxels a value of 0.7, 10\% of these white 
matter voxels were randomly selected to be assigned a value of 0.6 in the simulated control 
template and a value of 0.3 in the simulated patient template.  
Note that these values were arbitrarily selected to be more or less in the range of typical
FA values.  Gaussian noise (mean = 0.0, standard deviation = 0.1) was added to the white
matter template followed by a Gaussian smoothing (standard deviation = 1.5 mm) shown in Fig. \ref{fig:simulated_data}(b).  A simulated control cohort was created consisting of 20 subjects.  
Starting with the simulated control template, each of the 20 subjects is created by adding
Gaussian noise (mean = 0.0, standard deviation = 0.1) followed by 
Gaussian smoothing (standard deviation = 1.5 mm).  Thus, each subject is differentiated only by a unique random draw from the noise distribution and there is no real shape difference between subjects.  A simulated patient cohort of 20 subjects
was created similarly using the simulated patient template.  A sample control subject and sample
patient subject are given, respectively, in Fig. \ref{fig:simulated_data}(c) and (d).

\begin{figure*}
\begin{center}
\begin{tabular}{c}
%  \includegraphics[width=180mm]{simulatedTesting.pdf}
  \includegraphics[width=130mm]{simulatedTesting.pdf}
\end{tabular}
\caption{Student's $t$-test and the nonparametric Wilcoxon rank-sum test results (one-tailed, $\alpha = 0.95$) for the simulated experiment adjusted for multiple comparisons (FDR).
The true significance curves are drawn in black.  Elevated statistical power is seen with all 4 
         metrics for both tests although the effect is less in the more robust Wilcoxon results. Not shown are the significance curves for the Brunner-Munzel testing \citep{Rorden2007} which showed similar effects to the results on the left.
        }
\label{fig:simulated_testing}
\end{center}        
\end{figure*}

@@

To illustrate the effect of similarity metric on statistical significance, we performed ANTs registration \citep{Avants2011} of each subject to the template
(Fig. \ref{fig:simulated_data}(b)) using cross correlation (CC), mutual information (MI), 
Demons \citep{Thirion1998}, and sum of squared differences (SSD). A representative deformed image
using the SSD metric and its corresponding deformed grid is given in 
Fig. \ref{fig:simulated_data_warped}. 
The resulting statistical significance curves are plotted with the true significance curves in Fig. 
\ref{fig:simulated_testing}.

@@

Common similarity metric functions can therefore cause false positives by minimizing local intensity 
variance between the subjects and template which explicitly maximizes statistical
testing results in contrast to maximizing anatomical alignment.  This effect is strongest in 
the SSD and Demons metrics and weakest in the MI and CC metrics likely due
to the degree of correspondence to Eqn. (\ref{eq:variance}).   This is also supported by contrasting
the $t$-test results given in Fig. \ref{fig:simulated_testing}(a) with the non-parametric Wilcoxon
rank-sum testing (Fig. \ref{fig:simulated_testing}(b)).  The rank test does not explicitly use variance in its assessment and thus exhibits far less statistical exaggeration even though it offers roughly 95\%
the power of the $t$-test assuming certain assumptions are met \citep{Siegel1988,Rorden2007}.

Although various processing choices mitigate these effects, such as Gaussian smoothing, nonparametric testing, and the use of more robust similarity metrics, such choices are often made on an ad hoc basis.  
In contrast, in this work we strongly advocate the general application
%A more principled approach is the use of an anatomical, i.e. T1-weighted, unbiased shape/intensity 
%template constructed from the population images.  DTI-derived scalar images can then be mapped 
%to their corresponding T1-weighted image in a highly constrained manner and subsequently 
%mapped to the template via nonrigid registration of the T1-weighted image.  This minimizes 
%possible complications associated with direct FA-to-FA image registration.
%We recognize that using the anatomical images in the alignment process has been previously proposed (in fact, such a suggestion occurs in the {\it Future directions} section of the original TBSS paper).  
of a high performance \citep{Klein2009}, nonrigid registration algorithm \citep{Avants2011}, which forms the underlying machinery for an optimal template construction algorithm \citep{Avants2010}, in facilitating alignment between DTI-derived scalar images for cross-population studies.  The key point, here, is that the FA images are mapped to the common reference space via intra-subject DT-to-T1 mapping followed by T1-to-T1 alignment.  The T1-to-T1 step is (statistically) independent of the FA values which are the ultimate statistical target of hypothesis testing in FA population studies.  Below, we show that this same pattern of effects is reproduced in real data and detail our proposed solution.

\begin{figure*}
\begin{center}
\begin{tabular}{c}
%  \includegraphics[width=170mm]{workflow.pdf}
  \includegraphics[width=130mm]{workflow.pdf}
\end{tabular}
\caption{Possible ANTs-based TBSS workflow using SyGN.  In contrast to standard TBSS in which an individual subject's FA is registered to the FMRIB58 template
(transform represented with the blue arrow), our proposed modification first constructs an ANTs optimal anatomical template as illustrated in the center of the figure.  Each FA image is then rigidly mapped, with distortion correction, to the individual T1 using the white matter mask as shown on the left.  The individual T1-weighted image is then nonrigidly registered to the anatomical template.  Optionally, one can register the T1 template to the MNI152 template which resides in the same space as the FMRIB58 template.  Thus, the composite transform, carefully constructed without any FA-to-FA image registration, can take an individual subject's FA map (or other DTI-derived measure) to the desired standardized coordinate system.    }
\label{fig:workflow}
\end{center}
\end{figure*}

@@

%
%
%
%It was precisely due to these concerns that the TBSS framework was developed constituting an entire processing pipeline spanning from the reconstruction of the DTI tensorial data
%%to the calculation of the corresponding DTI-derived scalar measures (typically fractional anisotropy (FA)), to the nonrigid alignment of the population to a standard coordinate system and skeletonization, and, finally, 
%through to the statistical analysis for discerning differences across populations.  Major intermediate steps include the calculation of the corresponding DTI-derived scalar measures (typically fractional anisotropy (FA) \citep{Basser1996}), nonrigid alignment of the population to a standard coordinate system, and subsequent ``skeletonization'' of the mean of the aligned FA maps.  The alignment to a standard space is performed in such a way to remove the smoothing step associated with SPM.  Skeletonization and its projection across all registered DTI-derived scalar images is meant to accommodate the possibility of imperfectly aligned data.  However, as shown in a recent study \citep{Zalesky2011}, although the alignment error is reduced with the skeleton projection step, misalignments still remain post-projection.  Added to this is the potential confounding effect of minimizing data differences via nonrigid registration which are potentially the precise differences to be statistically quantified.

@@

%A crucial component of TBSS is the alignment of the DTI-derived scalar images \citep{Zalesky2011}. 
%
%Voxel-based morphometry vs. tractography vs. tract-specific analysis
%
%Talk about {\em skeleton projection} and how our work is different from \citep{Zalesky2011}.
%Also talk about the advantages over building a dti-based template as discussed in 
%\citep{Hecke2010}.

@@

\subsection{Anatomical Image Template Construction}
In general, we normalize images to to a standard coordinate system to
reduce intersubject variability in population studies.  Various
approaches exist for determining the normalized space such as selection
of a pre-existing template based on a single subject, e.g. the Talairach
atlas \citep{Talairach1988}, or a publicly available averaged group of
subjects, e.g. the MNI \citep{Collins1994} or ICBM \citep{Mazziotta1995}
templates.  One challenge with standard templates is that they may inadvertantly bias one's results by enabling better normalization of subjects to which the template is more similar.  This issue is exacerbated when dealing with populations that have high variance (e.g. due to disease) and/or when one's normalization method is low-dimensional (not flexible enough to capture large shape differences).

Population-specific templates (or optimal templates) alleviate some of
these issues by deriving a most representative image from the population
\citep{Good2001}.  Large deformation registration algorithms also reduce
this confound by being less sensitive to the deformation distance
between subject and target.  Some approaches combine both advantages,
for instance, the diffeomorphic approach of Joshi et al. employs the
SSD metric and a shape distance to bring the subject group of images
into alignment \citep{Joshi2004}.  Variants
include extension to multiple modalities \citep{Lorenzen2006} and small deformations
\citep{Geng2009}.  These approaches iteratively minimize group difference in ``congealing''
towards a representative image template \citep{Learned-Miller2006}.

In contrast, Avants et al. explicitly model the geometric component of the 
normalized space during optimization.  Coupling the intrinsic symmetry of 
SyN pairwise registration \citep{Avants2011} and an
optimized sharpening/averaging of the template appearance, Symmetric Group
Normalization (SyGN) is a powerful framework for producing optimal population-specific
templates \citep{Avants2010} with arbitrary similarity metric choice. Given a set of representative images, 
$\{I_1, I_2, \ldots, I_M\}$, optimization involves finding the set of paired
diffeomorphic transformations, $\left\{\left(\phi^1_1, \phi^1_2\right), 
\left(\phi^2_1, \phi^2_2\right), \ldots, \left(\phi^M_1, \phi^M_2\right) \right\}$,
the optimal template appearance, $J^*$, and corresponding coordinate system, $\psi(\mathbf{x})$,
which minimize the following cost function:
\begin{align}
  \sum_{m=1}^M \left[
    D\left( \psi(\mathbf{x}), \phi^m_1(\mathbf{x}, 1) \right) + 
    \Pi \left( I_m\left(\phi^m_2(\mathbf{x}, 0.5) \right), J^*\left(\phi^m_1(\mathbf{x}, 0.5) \right) \right)
    \right]
\end{align}
where $D$ is the diffeomorphic shape distance, 
\begin{align}
  D\left(\phi(\mathbf{x}, 0), \phi(\mathbf{x}, 1)\right) = \int_{0}^1 \| v(\mathbf{x}, t) \|_L dt, 
\end{align}
dependent upon the choice of the linear operator, $L$, and
$v$ is the diffeomorphism-generating velocity field, 
\begin{align}
  v\left(\phi(\mathbf{x}, t), t \right) = \frac{d\phi(\mathbf{x}, t)}{dt},\,\,\, \phi(\mathbf{x}, 0) = \mathbf{x}.
\end{align}
$\Pi$ is the choice of similarity metric, often cross-correlation \citep{Avants2008a}, calculated in the 
virtual domain midway between each individual image and the current estimate of the template.

@@

\begin{enumerate}
  \item Build the anatomical template from  the population sample or sample subset using SyGN as described in
  the previous section.
  \item If only using a sample subset for template creation, register the remaining
        anatomical data to the template and save the resulting transformation for each         
        subject indexed by $i$ (i.e. $\mathcal{T}^i_{template}$).
  \item Find the optimal transformation from each
        subject's FA to the anatomical template:
  \begin{enumerate}
  \item For each anatomical image, segment the white matter (in ANTs, one can use N4 bias correction \citep{Tustison2010} and Atropos $n$-tissue segmentation \citep{Avants2011a}).
  \item Rigidly transform each subject's average DWI to the bias-corrected anatomical image ($\mathcal{R}^i_{dwi}$).
  \item Perform distortion correction on the FA to anatomical mapping 
        by localized nonrigid registration
        of rigidly transformed FA map ($=\mathcal{R}^i_{dwi} \mathrm{FA}$) using
        the white matter mask obtained in step 3(a) ($\mathcal{T}^i_{FA}$).
  \item For each subject, concatenate the transforms which maps each
        subject's FA to the template ($\mathcal{T}^i_{total} = \mathcal{T}^i_{template} \circ 
        \mathcal{T}^i_{FA} \circ \mathcal{R}^i_{dwi} \mathrm{FA}$).
  \item Warp each subject's FA map to the template using $\mathcal{T}^i_{total}$.  
  Equivalently, one could warp the mean diffusion (MD), axial diffusion (AD) or any 
  other scalar derivable from DTI to the template.  
%  One other detail is that we prefer to warp the subject's DTI to 
%  the template and then, in template space, derive the scalar images of interest.  
  \end{enumerate}      
%  \item Perform the remaining portions of the standard TBSS pipeline:
%    \begin{enumerate}
%      \item Create the mean FA image from all template-aligned FA images.
%      \item Produce the white matter skeleton from the mean FA image.
%      \item Create skeleton distance map and project all FA data onto the skeleton.
%      \item Run the statistical analysis.
%    \end{enumerate}
\end{enumerate}
The next section compares this framework to FA-based normalization strategies with respect to both a patient-control study and a study of gender effects in a control population.

\subsection{Imaging Data}
Diffusion tensor data associated with each of the following data sets were reconstructed from the diffusion weighted sequences using Camino%
\footnote{
http://camino.org.uk/
}---an open source toolkit for diffusion MRI processing and analysis \citep{Cook2006} in combination with ANTs-based registration for motion correction of the DTI sequence.
